---
title: "Vacation Recommendation"
author: "Human Learning"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE
)
```

```{r hidden-functions, include=FALSE}
#########
# 1. K-Nearest Neighbors
###########################

knn = function(X, Y, myx, k, type=2, ztype="z-score", weight=NULL) {
  n = nrow(X); p = ncol(X)
  
  if(ztype == 'z-score') {
    Xs = scale(X)
    myx_s = (myx - colMeans(X)) / apply(X, 2, sd)
  } else if(ztype == 'min-max') {
    xmin = apply(X, 2, min)
    xmax = apply(X, 2, max)
    rng = xmax - xmin
    Xs = sweep(sweep(X, 2, xmin, "-"), 2, rng, "/")
    myx_s = (myx - xmin) / rng
  } else {
    Xs = X; myx_s = myx
  }
  
  diff = Xs - matrix(myx_s, nrow=n, ncol=p, byrow=TRUE)
  
  if (!is.null(weight)) {
    if (length(weight)==p) diff = sweep(diff, 2, weight, "*")
    else diff = diff * matrix(weight, nrow=n, ncol=p, byrow=FALSE)
  }
  
  if(type==1)      dist = rowSums(abs(diff))
  else if(type==2) dist = sqrt(rowSums(diff^2))
  else if(type=="infty") dist = apply(abs(diff), 1, max)
  else dist = (rowSums(abs(diff))^type)^(1/type)
  
  Y = factor(Y)
  idx = order(dist)[1:k]
  ylabel = Y[idx]
  
  tab = table(ylabel)
  distri_table = data.frame(
    class = names(tab),
    count = as.integer(tab),
    probs = as.integer(tab)/k
  )
  
  list(
    k_nearest_labels = ylabel,
    distribution_table = distri_table,
    prob = max(distri_table$probs),
    yhat = names(tab)[which.max(tab)]
  )
}

myknnacc = function(Xtrain, Ytrain, Xtest, Ytest, k, type=2, ztype="z-score", weight=NULL) {
  ntest = nrow(Xtest)
  preds = sapply(1:ntest, function(i) knn(Xtrain, Ytrain, Xtest[i,], k, type, ztype, weight)$yhat)
  list(
    yhat = preds,
    accuracy = mean(preds == Ytest)
  )
}


###########################
# 2. K-Means Clustering
###########################

kmc = function(X, k, init='Forgy', iter='Lloyd', seed){
  set.seed(seed)
  n = nrow(X); p = ncol(X)
  
  if (init=="Forgy") {
    centroids = X[sample.int(n,k),]
  } else stop("Only Forgy used here for simplicity.")
  
  clusters = rep(0,n)
  
  repeat {
    dist = sapply(1:k, function(j) rowSums((X - matrix(centroids[j,],n,p,TRUE))^2))
    new_clusters = max.col(-dist)
    if(all(new_clusters==clusters)) break
    clusters = new_clusters
    centroids = sapply(1:k, function(j) colMeans(X[clusters==j,,drop=FALSE]))
    centroids = t(centroids)
  }
  
  nearest_cids = function(myx) {
    d = rowSums((centroids - matrix(myx,k,p,TRUE))^2)
    which.min(d)
  }
  
  list(
    cids = clusters,
    final_centroids = centroids,
    nearest_cids = nearest_cids
  )
}
```

# I. Introduction

Choosing an ideal travel destination is a complicated decision making process. Travelers must balance personal preferences, such as climate, travel style, and budget, with contextual factors like trip duration, companionship, and geographic constraints. Therefore, we developed a machine learning driven vacation recommendation engine that predicts destination ratings and produces personalized Top-5 travel suggestions.

Our system integrates a set of user level features (demographics, preferences, behavioral patterns) and location-level characteristics (climate zone, city type, average temperature, and environmental attributes). Past user ratings are parsed and transformed into an expanded training dataset, enabling the model to learn relationships between user profiles and detination desirability.

To model these relationships, we use a diverse ensemble of predictive methods, including: - K-Neareast Neighbors (KNN) - Decision Tree - Random Forest - Support Vector Machine (SVM) - K-means clustering - LightGBM gradient boosting regression - XGBoost Regression

Rather than relying on a single model, we combine all using a stacking meta-learner, trained through cross-validation predictions. This meta-model, implemented as a linear regression, automatically learns the optimal weighting of each base model, yielding a more stable and accurate final rating predictor compared to soft voting or individual models.

The system supports two recommendation scenarios:

1.  Destination ranking without user input, based only on aggregated historical ratings.
2.  Personalized Top-5 recommendations, where each destination's predicted rating is computed using stacking ensemble. Optional user-specified filters allow further customization.

Overall, this project delivers a practical, data-driven recommendation system that generates personalized travel suggestions using a stacked ensemble of machine learning models. The combination of diverse predictors and careful feature engineering allows the system to produce reliable and adaptable destination rankings for a wide range of travelers.

## Dataset

We use two datasets: a user dataset containing individual characteristics and past travel ratings, and a destination dataset describing attributes of each location.

```{R}
users = read.csv('src/Vacation_UsersData.csv')
dests = read.csv('src/Vacation_DestinationData.csv')
```

Key variables include:

User Dataset - Age, ParentalEducationYears: demographic information - CEOAnnualSalary, Budget: socioeconomic and financial factors - IdealTripDurationDays: preferred trip length - CurrentLatitude, CurrentLongtitude: geographic context - ClimatePreference, TravelStylePreference, TravelCompanionPreference: categorical travel preferences - FavoriteColor, LocalSportsTeamRecord: additional preference indicators - PastTripsWithRatings: list of visited destinations and corresponding ratings.

Destination Dataset - LocationID: identifier for each destination - ClimateZone: environmental category of the destination - CityType: type of urban environment - AvgTemperature_C: average local temperature - Latitude & Longtitude

# II. Methods

We begin by importing all the functions (will be attached in the back) and packages used in this project.

```{R}
source('src/knn.R')
source('src/kmc.R')
library(rpart)
library(randomForest)
library(e1071)
library(xgboost)
library(lightgbm)
```

Note: For the MLR and SVM parts, we use package implementations instead of our own course functions. Because the one-hot encoded design matrix is very sparse and $X^\top X$ is not invertible, our custom MLR function based on $(X^\top X)^{-1}$ cannot be used. Therefore, for both MLR and SVM we rely on the corresponding package implementations, which are numerically stable under high dimensional and sigular design matrices.

User preference variable are converted to factors so they can be properly encoded in the design matrix during model fitting.

```{R}

users$ClimatePreference = factor(users$ClimatePreference)
users$TravelStylePreference = factor(users$TravelStylePreference)
users$TravelCompanionPreference = factor(users$TravelCompanionPreference)
users$FavoriteColor = factor(users$FavoriteColor)
users$LocalSportsTeamRecord = factor(users$LocalSportsTeamRecord)
```

## 1. Data Preprocession

We split the user dataset into 65% training and 35% testing with a fixed random seed for reproducibility. These subsets are used throughout model training and evaluation.

```{R}
set.seed(1234)
n = nrow(users)
ptrain = 0.65
ntrain = floor(n*ptrain)
idx = seq(1, n, 1)
itrain = sample(idx, ntrain, replace=F)
itest = setdiff(idx, itrain)
Xtrain = users[itrain,]
Xtest = users[itest,]
```

Each user's past travel history is stored as a string containing destination IDs and corresponding ratings. We parse this field into a structured format containing one row per (user, destination) pair. We then expand the dataset so that each user appears once per visited destination and merge these records with destination attributes. We remove identifier and non-predictive variable, then construct a fully encoded design matrix using model.matrix. The final training features (Xtrain) and response vector (Ytrain) are extracted for model fitting. We remove identifier and non-predictive variable, then construct a fully encoded design matrix using model.matrix. The final training features (Xtrain) and response vector (Ytrain) are extracted for model fitting.

```{R}
preprocess = function(Xtrain, dests) {
  parse_trips = function(trips) {
    parts = unlist(strsplit(trips, ", "))
    
    loc = c()
    rat = c()
    for (p in parts) {
      reg = regexec("Destination_([0-9]+)\\s*-\\s*([0-9]+)", p)
      output = unlist(regmatches(p, reg))
      loc = c(loc, as.integer(output[2]))
      rat = c(rat, as.integer(output[3]))
    }
    return(data.frame(LocationID = loc, Rating = rat))
  }
  
  Xtrain_list = lapply(1:nrow(Xtrain), function(i) {
    trips = parse_trips(Xtrain$PastTripsWithRatings[i])
    base = Xtrain[i,]
    base = base[rep(1, nrow(trips)), , drop=F]
    return(cbind(base, trips))
  })
  
  Xtrain_all = do.call(rbind, Xtrain_list)
  Xtrain_merge = merge(Xtrain_all, dests, by = "LocationID")
  
  drop_cols = c("UserID", "PastTripsWithRatings", "Rating", "CityName", "LocationID")
  feature_cols = setdiff(names(Xtrain_merge), drop_cols)
  
  Xtrain = model.matrix(~ . -1, data = Xtrain_merge[, feature_cols, drop=F])
  Ytrain = Xtrain_merge$Rating
  return(list(Xtrain = Xtrain, Ytrain = Ytrain, 
              Xtrain_all = Xtrain_all, feature_cols = feature_cols))
}

output = preprocess(Xtrain, dests)
feature_cols = output$feature_cols
Xtrain_all = output$Xtrain_all
Xtrain = output$Xtrain
Ytrain = output$Ytrain
```

Because the set of one-hot-encoded columns must remain consistent across training and testing, we define a helper function to align any new matrix to the training column stucture.

```{R}
train_cols = colnames(Xtrain)
align_matrix = function(X_new, train_cols) {
  missing_cols = setdiff(train_cols, colnames(X_new))
  if (length(missing_cols) > 0) {
    add = matrix(0, nrow = nrow(X_new), ncol = length(missing_cols))
    colnames(add) = missing_cols
    X_new = cbind(X_new, add)
  }
  return(X_new[, train_cols, drop=F])
}
```

## 2. Train the 7 base models

To capture different aspects of the relationship between user-destination features and ratings, we fit a diverse set of regression models on the same design matrix Xtrain and response vector Ytrain. The function below trains all seven base models and returns them in a single list object:

```{R}
## All hyperparameters used below (e.g., k_knn = 17, k_kmc = 3, ntree_rf = 1000)
## are the final values selected from our hyperparameter tuning procedures shown
## in Section IV (Sidenote).
train_all_rating = function(Xtrain, Ytrain, k_knn = 17, k_kmc = 3,
                            ntree_rf = 1000, seed = 1234) {
  
  # 1) KNN regression: 
  knn_model = list(X = Xtrain, Y = Ytrain, k = k_knn)
  
  # 2) DT regression
  dt_df = data.frame(Rating = Ytrain, Xtrain)
  dt_model = rpart(Rating ~ ., data = dt_df, method = "anova",
                   control = rpart.control(cp = 0.001, minsplit = 10))
  
  # 3) RF regression
  rf_model = randomForest(x = Xtrain, y = Ytrain,
                          ntree = ntree_rf,
                          mtry = floor(sqrt(ncol(Xtrain))))
  
  # 4) SVM regression
  svm_model = e1071::svm(x = Xtrain, y = Ytrain, 
                         type  = "eps-regression", kernel = "radial")
  
  # 5) KMC
  kmc_model_raw = kmc(Xtrain, k = k_kmc, init = "Forgy", iter = "Lloyd", seed = seed)
  mean_rate = tapply(Ytrain, kmc_model_raw$cids, mean)
  kmc_model = list(kmc = kmc_model_raw, mean_rate = mean_rate)
  
  # 6) LightGBM regression
  lgbm_train = lgb.Dataset(data = Xtrain, label = Ytrain)
  lgbm_params = list(objective        = "regression",
                     metric           = "rmse",
                     learning_rate    = 0.05,
                     num_leaves       = 31,
                     feature_fraction = 0.8,
                     bagging_fraction = 0.8,
                     bagging_freq     = 1)
  
  lgbm_model = lgb.train(params  = lgbm_params,
                         data    = lgbm_train,
                         nrounds = 400,
                         verbose = -1)
  
  # 7) XGBoost regression
  dtrain = xgb.DMatrix(data = Xtrain, label = Ytrain)
  params = list(objective        = "reg:squarederror",
                eval_metric      = "rmse",
                max_depth        = 4,
                eta              = 0.05,
                subsample        = 0.8,
                colsample_bytree = 0.8)
  xgb_model = xgb.train(params = params, data = dtrain, nrounds = 400, verbose = 0)
  
  return(list("knn"   = knn_model,
              "dt"    = dt_model,
              "rf"    = rf_model,
              "svm"   = svm_model,
              "kmc"   = kmc_model,
              "lgbm"  = lgbm_model,
              "xgb"   = xgb_model))
}

models = train_all_rating(Xtrain, Ytrain)
```

## 3. Per-model rating predictors

After fitting the seven base models, we define a corresponding prediction function for each one. These functions take a new design matrix Xnew and return the model specific estimated ratings:

### 1) KNN average neighbor ratings

Computes the average rating among the k nearest neighbors of each new observation in feature space.

```{R}
knn_rating = function(model, Xnew) {
  n_new = nrow(Xnew)
  ratings = numeric(n_new)
  
  for (i in 1:n_new) {
    KNN = knn(model$X, model$Y, Xnew[i, ], k = model$k, type = 2, ztype = "z-score")
    labs = as.numeric(KNN$k_nearest_labels)
    ratings[i] = mean(labs)
  }
  return(ratings)
}
```

### 2) Decision Tree regression

Uses the fitted regression tree to assign each observation to a leaf node and return the node’s predicted value.

```{R}
dt_rating = function(model, Xnew) {
  as.numeric(predict(model, newdata = as.data.frame(Xnew)))
}
```

### 3) Random Forest regression

Aggregates predictions across all trees in the ensemble, getting a stable estimate.

```{R}
rf_rating = function(model, Xnew) {
  as.numeric(predict(model, newdata = Xnew))
}
```

### 4) SVM regression

Uses a SVM model to estimate ratings based on patterns it learned from the training data.

```{R}
svm_rating = function(model, Xnew) {
  as.numeric(predict(model, newdata = Xnew))
}
```

### 5) KMC cluster mean rating

Assigns each new point to the nearest cluster and uses that cluster’s mean rating as the prediction.

```{R}
kmc_rating = function(model, Xnew) {
  km = model$kmc
  mr = model$mean_rate
  n_new = nrow(Xnew)
  ratings = numeric(n_new)
  
  for (i in 1:n_new) {
    cid = km$nearest_cids(Xnew[i, ])
    v = mr[as.character(cid)]
    ratings[i] = ifelse(is.na(v), mean(mr, na.rm=T), v)
  }
  return(ratings)
}
```

### 6) LightGBM regression

Applies the trained gradient boosting model to obtain fast, high-accuracy predictions.

```{R}
lgbm_rating = function(model, Xnew) {
  as.numeric(predict(model, Xnew))
}
```

### 7) XGB regression

Uses the trained XGBoost model to predict ratings based on many small decision trees working together.

```{R}
xgb_rating = function(model, Xnew) {
  as.numeric(predict(model, xgb.DMatrix(Xnew)))
}
```

These functions standardize the output across models so that each returns a numeric predicted rating for any new input matrix.

## 4. Meta-Learner for Stacking by using MLR

In our stacking framework, the seven base models are trained in parallel and each produces its own predicted rating. Instead of selecting a single model or averaging their outputs, we train a second-stage model using a simple linear regression (MLR), using these predictions as inputs.

This meta-learner effectively learns how much weight to assign to each base model. The coefficients of the MLR represent these learned weights, allowing the ensemble to automatically emphasize models that perform better and down weight those that are less accurate.

In short, the stacking layer refines the final prediction by optimally combining all seven model outputs, leading to improved accuracy and robustness.

```{R}
base_models = c("knn", "dt", "rf", "svm", "kmc", "lgbm", "xgb")
M = length(base_models)

build_meta = function(Xtrain, Ytrain, Kfolds = 5, seed = 1234) {
  set.seed(seed)
  n = nrow(Xtrain)
  folds = sample(rep(1:Kfolds, length.out = n))
  
  P = matrix(NA, nrow = n, ncol = M)
  colnames(P) = base_models
  
  for (k in 1:Kfolds) {
    idx_tr  = which(folds != k)
    idx_val = which(folds == k)
    
    model_k = train_all_rating(Xtrain[idx_tr, , drop=F], Ytrain[idx_tr], seed = seed)
    
    P[idx_val, "knn"] = knn_rating(model_k$knn, Xtrain[idx_val, , drop=F])
    P[idx_val, "dt"] = dt_rating(model_k$dt, Xtrain[idx_val, , drop=F])
    P[idx_val, "rf"] = rf_rating(model_k$rf, Xtrain[idx_val, , drop=F])
    P[idx_val, "svm"] = svm_rating(model_k$svm, Xtrain[idx_val, , drop=F])
    P[idx_val, "kmc"] = kmc_rating(model_k$kmc, Xtrain[idx_val, , drop=F])
    P[idx_val, "lgbm"] = lgbm_rating(model_k$lgbm, Xtrain[idx_val, , drop=F])
    P[idx_val, "xgb"] = xgb_rating(model_k$xgb, Xtrain[idx_val, , drop=F])
  }
  
  return(list(P = P, Y = Ytrain))
}

meta_data = build_meta(Xtrain, Ytrain, Kfolds = 5, seed = 1234)
meta_df = data.frame(Rating = meta_data$Y, meta_data$P)
head(meta_df, 20)

stack = lm(Rating ~ . -1, data = meta_df)
stack$coef
```

## 5. Models Ensemble: Stacking

For each new input, we first collect predictions from all seven base models. These values are then passed to the meta-model, which combines them using the learned weights to produce the final rating.

```{R}
stack_predict_rating = function(models, stack_model, Xnew) {
  comps = data.frame(knn = knn_rating(models$knn, Xnew),
                     dt  = dt_rating(models$dt, Xnew),
                     rf  = rf_rating(models$rf, Xnew),
                     svm = svm_rating(models$svm, Xnew),
                     kmc = kmc_rating(models$kmc, Xnew),
                     lgbm  = lgbm_rating(models$lgbm, Xnew),
                     xgb = xgb_rating(models$xgb, Xnew))
  
  return(as.numeric(predict(stack_model, newdata = comps)))
}
```

# III. Recommendation Scenerios

Here we show how the recommendation system works in two different situations.

## Case I: No user information

When no user preferences are available, the system relies solely on historical data. We compute the average rating received by each destination in the training set and recommend the Top-5 destinations with the highest mean scores.

```{R}
top5_no_info = function(Xtrain_all, dests, topk = 5) {
  avg = aggregate(Rating ~ LocationID, data = Xtrain_all, FUN = mean)
  avg = avg[order(-avg$Rating),]
  top = head(avg, topk)
  result = merge(top, dests, by = "LocationID")
  return(result[order(-result$Rating),])
}

top5_no_info(Xtrain_all, dests, topk = 5)
```

## Case II: User information available

When user data are provided, the system generates personalized destination recommendations. For each new user, we first cross join the user with all destinations to create a full user–destination grid. Every row in this grid represents “this user visiting this destination,” including both user features and destination features. The grid is then converted into the exact model, matrix structure used during training.

```{R}
### Case II: Have user info
prepare_Xnew = function(Xnew, dests, models, stack) {
  Xnew$id = seq_len(nrow(Xnew))
  grid0 = merge(Xnew, dests, by = NULL)
  
  prepare_grid = function(grid) {
    Xg = model.matrix(~ . -1, data = grid[, feature_cols, drop=F])
    Xg = align_matrix(Xg, train_cols)
    list(grid = grid, Xg = Xg)
  }
  
  prep = prepare_grid(grid0)
  grid = prep$grid
  Xg = prep$Xg
  
  grid$PredRating = stack_predict_rating(models, stack, Xg)
  out = split(grid, grid$id)
  return(out)
}
```

Next, each base model produces a predicted rating for every destination. The stacking meta-learner then combines these predictions into a single final rating. Users may also apply optional filters, such as climate preference, city type, or temperature range to restrict the candidate destinations before ranking.

```{R}
recommend_topk_rating = function(Xnew, dests, topk = 5, models, stack, 
                                 climate_choice = NULL, city_choice = NULL, temp_choice = NULL) {
  out = prepare_Xnew(Xnew, dests, models, stack)
  result = lapply(out, function(df) {
    if (!is.null(climate_choice)) {
      df = df[which(df$ClimateZone == climate_choice),]
    }
    if (!is.null(city_choice)) {
      df = df[which(df$CityType == city_choice),]
    }
    if (!is.null(temp_choice)) {
      reg = gregexpr("-?\\d+\\.?\\d*", temp_choice)
      nums = as.numeric(unlist(regmatches(temp_choice, reg)))
      df = df[which(df$AvgTemperature_C > nums[1] & 
                    df$AvgTemperature_C <= nums[2]),]
    }
    df = df[order(df$PredRating, decreasing=T), , drop=F]
    return(head(data.frame(LocationID = df$LocationID, PredRating = df$PredRating), topk))
  })
  return(result)
}
```

Finally, the top-ranked destinations are merged back with their metadata (city type, climate zone, etc.) and packaged together with the corresponding user. This provides an interpretable output structure: for each user, the system reports the 5 destinations with the highest predicted ratings according to the ensemble model.

```{R}
top5_have_info = function(Xnew, dests, topk = 5, models, stack,
                          climate_choice = NULL, city_choice = NULL, temp_choice = NULL) {
  model = recommend_topk_rating(Xnew, dests, topk, models, stack,
                                climate_choice, city_choice, temp_choice)
  
  loc_list = vector("list", length = nrow(Xnew))
  for (i in seq_len(nrow(Xnew))) {
    top5 = model[[i]]
    merged  = merge(top5, dests, by = "LocationID")
    loc_list[[i]]  = merged[order(merged$PredRating, decreasing=T), ]
  }
  
  user_list <- split(Xnew, seq_len(nrow(Xnew)))
  result <- Map(function(user, dests) {
      list(user = user, dests = dests)
    }, user_list, loc_list)
  
  return(result)
}
```

### Examples

#### Example 1 (recommend based on test set)

We first demonstrate how the system performs personalized recommendations for several users drawn directly from the test set. For each user profile, the model evaluates all destinations, predicts a rating through the stacked ensemble, and returns the Top-5 destinations with the highest predicted scores.

```{R}
new_users = Xtest[1:5,]
top5_have_info(Xnew = new_users, dests, topk = 5, models, stack = stack)
```

#### Example 2 (interaction to get user info)

Note: The follow-up filtering step uses menu(), which is only available in interactive R sessions. When knitting this document, default choices are used to avoid interrupting the render process.

The system can also generate recommendations for a user who provides their information interactively. After collecting numeric and categorical inputs, the profile is encoded to match the training feature space and passed into the stacked model to produce personalized Top-5 recommendations.

```{R}
top5_ask_info = function(dests, topk = 5, models, stack) {
  
  # ---------- Numeric inputs ----------
  if (interactive()) {
    age     = readline("Please enter your age: ")
    parent  = readline("Please enter your parental education years: ")
    ceo     = readline("Please enter your CEO annual salary: ")
    budget  = readline("Please enter your budget: ")
    duration= readline("Please enter your ideal trip duration days: ")
    lat     = readline("Please enter your current latitude: ")
    long    = readline("Please enter your current longitude: ")
  } else {
    # defaults when knitting
    age = parent = ceo = budget = duration = lat = long = 0  
  }
  
  # ---------- Climate ----------
  if (interactive()) {
    climate_idx = menu(c("Cold","Mixed","Temperate","Warm"), 
                       title = "Please choose climate:")
  } else {
    climate_idx = 3   # default = Temperate
  }
  climate = c("Cold","Mixed","Temperate","Warm")[climate_idx]
  
  # ---------- Style ----------
  if (interactive()) {
    style_idx = menu(c("Adventure","Cultural","Luxury","Party","Relaxation"),
                     title = "Please choose travel style:")
  } else {
    style_idx = 2
  }
  style = c("Adventure","Cultural","Luxury","Party","Relaxation")[style_idx]
  
  # ---------- Companion ----------
  if (interactive()) {
    compan_idx = menu(c("Couples","Family","Friends","Solo","Work"),
                      title = "Please choose companion type:")
  } else {
    compan_idx = 1
  }
  compan = c("Couples","Family","Friends","Solo","Work")[compan_idx]
  
  # ---------- Color ----------
  if (interactive()) {
    color_idx = menu(c("Black","Blue","Green","Orange","Purple","Red","White","Yellow"),
                     title = "Please choose favorite color:")
  } else {
    color_idx = 1
  }
  color = c("Black","Blue","Green","Orange","Purple","Red","White","Yellow")[color_idx]
  
  # ---------- Sports team ----------
  if (interactive()) {
    sport_idx = menu(c("Losing","Neutral","Winning"),
                     title = "Local sports team record:")
  } else {
    sport_idx = 2   # Neutral
  }
  sport = c("Losing","Neutral","Winning")[sport_idx]
  
  
  new_users = data.frame(
    Age                     = as.numeric(age),
    ParentalEducationYears  = as.numeric(parent),
    CEOAnnualSalary         = as.numeric(ceo),
    Budget                  = as.numeric(budget),
    IdealTripDurationDays   = as.numeric(duration),
    CurrentLatitude         = as.numeric(lat),
    CurrentLongitude        = as.numeric(long),
    ClimatePreference       = climate,
    TravelStylePreference   = style,
    TravelCompanionPreference = compan,
    FavoriteColor           = color,
    LocalSportsTeamRecord   = sport
  )

  new_users$ClimatePreference       = factor(new_users$ClimatePreference, 
                                             levels = levels(users$ClimatePreference))
  new_users$TravelStylePreference   = factor(new_users$TravelStylePreference, 
                                             levels = levels(users$TravelStylePreference))
  new_users$TravelCompanionPreference = factor(new_users$TravelCompanionPreference, 
                                              levels = levels(users$TravelCompanionPreference))
  new_users$FavoriteColor           = factor(new_users$FavoriteColor, 
                                             levels = levels(users$FavoriteColor))
  new_users$LocalSportsTeamRecord   = factor(new_users$LocalSportsTeamRecord, 
                                             levels = levels(users$LocalSportsTeamRecord))
  
  top5_have_info(new_users, dests, topk, models, stack)
}

top5_ask_info(dests, topk = 5, models = models, stack = stack)

```

## Case III: Have user info (can ask follow up questions about destination info)

Beyond the initial personalized recommendations, the system allows optional follow-up filters, such as climate, city type, or temperature range to refine the results. If the user selects “I’m fine with all,” no additional filtering is applied.

```{R}
top5_followup_info = function(new_users, dests, topk = 5, models, stack) {
if (interactive()) {
  climate_options = menu(c("Dry","Temperate","Continental","Tropical","Polar",
                           "I'm fine with all"),
                         title = "Please choose one:")
} else {
  climate_options = 6  # skip interaction when knitting
}

climate_choice = if (climate_options == 6) NULL 
                 else c("Dry","Temperate","Continental","Tropical",
                        "Polar")[climate_options]


if (interactive()) {
  city_options = menu(c("Cultural","Resort","Beach","Urban","Natural",
                        "Adventure","I'm fine with all"),
                      title = "Please choose one:")
} else {
  city_options = 7
}

city_choice = if (city_options == 7) NULL
              else c("Cultural","Resort","Beach","Urban","Natural",
                     "Adventure")[city_options]


if (interactive()) {
  temp_options = menu(c("-10 < temp <= 0","0 < temp <= 10","10 < temp <= 20",
                        "20 < temp <= 33","I'm fine with all"),
                      title = "Please choose one:")
} else {
  temp_options = 5
}

temp_choice = if (temp_options == 5) NULL
              else c("-10 < temp <= 0","0 < temp <= 10","10 < temp <= 20",
                     "20 < temp <= 33")[temp_options]

  return(top5_have_info(new_users, dests, topk, models, stack, 
                        climate_choice, city_choice, temp_choice))
}

new_users = Xtest[1,]
top5_followup_info(new_users, dests, topk = 5, models, stack)
```

## Case IV: Partial Predictors (have same predictors as another team)

In this case, we consider a scenario where only a partial set of user information is available and shared with another team. The predictors include basic trip preferences, current location, climate preference, and past trips with ratings.

Using this limited information, we train rating models and generate the top 5 destination recommendations for a new user. This case evaluates whether reasonable recommendations can still be produced with commonly available user predictors.

```{r}
X1 = users$Budget
X2 = users$IdealTripDurationDays
X3 = users$CurrentLatitude
X4 = users$CurrentLongitude
X5 = users$ClimatePreference
X6 = users$PastTripsWithRatings
users4 = data.frame("Budget" = X1, 
                   "IdealTripDurationDays" = X2, 
                   "CurrentLatitude" = X3, 
                   "CurrentLongitude" = X4, 
                   "ClimatePreference" = X5,
                   "PastTripsWithRatings" = X6)
users4$ClimatePreference = factor(users4$ClimatePreference)

Xtrain4 = users[itrain,]
Xtest4 = users[itest,]
output4 = preprocess(Xtrain4, dests)
feature_cols4 = output4$feature_cols
Xtrain4 = output$Xtrain
Ytrain4 = output$Ytrain
train_cols4 = colnames(Xtrain4)

models4 = train_all_rating(Xtrain4, Ytrain4)
meta_data4 = build_meta(Xtrain4, Ytrain4, Kfolds = 5, seed = 1234)
meta_df4 = data.frame(Rating = meta_data4$Y, meta_data4$P)
stack4 = lm(Rating ~ . -1, data = meta_df4)

new_users = Xtest4[1,]
top5_have_info(new_users, dests, topk = 5, models4, stack)

```

# IV. Sidenote: Hyperparameter Tuning

This section shows how the final hyperparameters were selected via cross-validation. These tuned values are used in the main model training.

## Hyperparameter Tuning for KNN

```{R}
rmse = function(y, yhat) {
  sqrt(mean((y - yhat)^2))
}

tune_knn_k = function(X, y, k_grid = seq(3, 31, 2), Kfolds = 5, seed = 1234) {
  set.seed(seed)
  n = nrow(X)
  folds = sample(rep(1:Kfolds, length.out = n))
  results = data.frame(k = k_grid, RMSE = NA_real_)
  
  for (i in seq_along(k_grid)) {
    k_val = k_grid[i]
    rmse_vec = numeric(Kfolds)
    
    for (f in 1:Kfolds) {
      idx_tr  = which(folds != f)
      idx_val = which(folds == f)
      
      model_fold = list(X = X[idx_tr, , drop=F],
                        Y = y[idx_tr],
                        k = k_val)
      
      preds = knn_rating(model_fold, X[idx_val, , drop=F])
      rmse_vec[f] = rmse(y[idx_val], preds)
    }
    
    results$RMSE[i] = mean(rmse_vec)
  }
  
  best_idx = which.min(results$RMSE)
  return(list(best_k = results$k[best_idx], cv_results = results))
}

# knn_tuning = tune_knn_k(Xtrain, Ytrain)
# knn_tuning$best_k
# knn_tuning$cv_results
```

## Hyperparameter Tuning for KMC

```{R}
tune_kmc_k = function(X, y, k_grid = c(3, 5, 7, 10, 15), 
                      Kfolds = 5, seed = 1234) {
  set.seed(seed)
  n = nrow(X)
  folds = sample(rep(1:Kfolds, length.out = n))
  results = data.frame(k = k_grid, RMSE = NA_real_)
  
  for (i in seq_along(k_grid)) {
    k_val = k_grid[i]
    rmse_vec = numeric(Kfolds)
    
    for (f in 1:Kfolds) {
      idx_tr  = which(folds != f)
      idx_val = which(folds == f)
      
      kmc_model_raw = kmc(X[idx_tr, , drop=F], k = k_val,
                          init = "Forgy", iter = "Lloyd", seed = seed)
      mean_rate = tapply(y[idx_tr], kmc_model_raw$cids, mean)
      kmc_model = list(kmc = kmc_model_raw, mean_rate = mean_rate)
      
      preds = kmc_rating(kmc_model, X[idx_val, , drop=F])
      rmse_vec[f] = rmse(y[idx_val], preds)
    }
    
    results$RMSE[i] = mean(rmse_vec)
  }
  
  best_idx = which.min(results$RMSE)
  return(list(best_k = results$k[best_idx], cv_results = results))
}

# kmc_tuning = tune_kmc_k(Xtrain, Ytrain)
# kmc_tuning$best_k
# kmc_tuning$cv_results
```

## Hyperparameter Tuning for Random Forest

```{R}
tune_rf_ntree = function(X, y, ntree_grid = c(500, 1000, 1500, 2000),
                         Kfolds = 5, seed = 1234) {
  set.seed(seed)
  n = nrow(X)
  folds = sample(rep(1:Kfolds, length.out = n))
  results = data.frame(ntree = ntree_grid, RMSE = NA_real_)
  
  for (i in seq_along(ntree_grid)) {
    ntree_val = ntree_grid[i]
    rmse_vec = numeric(Kfolds)
    
    for (f in 1:Kfolds) {
      idx_tr  = which(folds != f)
      idx_val = which(folds == f)
      
      rf_model = randomForest(x = X[idx_tr, , drop=F],
                              y = y[idx_tr],
                              ntree = ntree_val,
                              mtry = floor(sqrt(ncol(X))))
      
      preds = predict(rf_model, newdata = X[idx_val, , drop=F])
      rmse_vec[f] = rmse(y[idx_val], preds)
    }
    
    results$RMSE[i] = mean(rmse_vec)
  }
  
  best_idx = which.min(results$RMSE)
  return(list(best_ntree = results$ntree[best_idx], cv_results = results))
}

# rf_tuning = tune_rf_ntree(Xtrain, Ytrain)
# rf_tuning$best_ntree
# rf_tuning$cv_results
```

# V Two Groups Comparison Analysis

## No-User-Info Comparison

**Our Team — No Info Top 5**

| Rank | Destination     | Notable Attributes         |
|------|-----------------|----------------------------|
| 1    | Destination_023 | Mild temp, natural scenery |
| 2    | Destination_064 | Warm, urban                |
| 3    | Destination_037 | Tropical resort            |
| 4    | Destination_039 | Coastal temperate          |
| 5    | Destination_033 | Temperate urban            |

**Other Team — No Info Top 5**

| Rank | Destination     | FinalScore Components                              |
|------------------|------------------|-------------------------------------|
| 1    | Destination_030 | High intrinsic quality (PCA), safe                 |
| 2    | Destination_023 | High intrinsic quality, same pick as our team’s #1 |
| 3    | Destination_018 | Strong PCA score                                   |
| 4    | Destination_022 | Good PCA and cluster quality                       |
| 5    | Destination_064 | Appears in both teams' lists                       |

Without user information, the two systems produce different results because they define “population preference” differently. Our team relies on historical average ratings, so our recommendations reflect what the average user genuinely enjoyed, leading to a more diverse set of destinations. The other team focuses on destination attributes like safety and cost, which prioritizes structurally “high-quality” places but may not represent what people actually prefer. In situations where no user information is available, our approach is arguably better because it is grounded directly in real user behavior, making it a more realistic estimate of population-level preferences.

## With-User-Info Comparison

**Our Team - With User Info**

| Rank | Destination | PredRating | ClimateZone | CityType |
|------|-------------|------------|-------------|----------|
| 1    | 064         | 7.9515     | Dry         | Urban    |
| 2    | 031         | 7.9429     | Dry         | Urban    |
| 3    | 056         | 7.9404     | Dry         | Cultural |
| 4    | 061         | 7.9259     | Dry         | Urban    |
| 5    | 001         | 7.9250     | Dry         | Cultural |

**Other Team - With User Info**

| Rank | Destination | CFScore | PopScore | FinalScore |
|------|-------------|---------|----------|------------|
| 1    | 044         | 1.00    | 0.7405   | 0.9222     |
| 2    | 013         | 0.875   | 0.6804   | 0.8166     |
| 3    | 027         | 0.7486  | 0.5348   | 0.6845     |
| 4    | 050         | 0.7568  | 0.3576   | 0.6370     |
| 5    | 069         | 0.8169  | 0.1804   | 0.6260     |

When both systems use the same user information (UserID = 1), our model personalizes recommendations more effectively. By using the user’s warm - climate preference, budget, and past ratings, our model consistently returns warm/dry destinations that match the user’s profile. The other team’s hybrid CF system relies mainly on similar users’ behavior and cluster popularity, which can overlook explicit preferences such as climate. As a result, our system provides recommendations that align more closely with what our testing user clearly likes. Therefore, for this specific user, our personalized results are more targeted and better justified.

# VI. Conclusion

This project built a personalized vacation recommendation system using a stacked ensemble of seven machine learning models. By combining diverse predictors and user–destination features, the model produces more stable and accurate rating estimates than any single method. The system supports both general and personalized Top-5 recommendations, with optional filters for user-specific preferences. Overall, it provides a flexible foundation for destination ranking and can be extended with additional data or user behavior signals.

# Appendix of Sourced Functions

## 1. K-Nearest Neighbors (KNN)

```{R}
knn = function(X, Y, myx, k, type=2, ztype="z-score", weight=NULL) {
  # X is a matrix, Y and myx are vectors
  n = nrow(X)
  p = ncol(X)
  
  # ------------------- Normalization --------------------
  if(ztype == 'z-score') {
    Xs = scale(X)
    myx_s = (myx - colMeans(X)) / apply(X, 2, sd)
  } else if(ztype == 'min-max') {
    xmin = apply(X, 2, min)
    xmax = apply(X, 2, max)
    rng = xmax - xmin
    Xs = sweep(sweep(X, 2, xmin, FUN = "-"), 2, rng, FUN = "/")
    myx_s = (myx - xmin) / rng
  } else if(ztype == 'none') {
    Xs = X
    myx_s = myx
  } else {
    stop("Unknown target space. Use 'z-score', 'min-max', or 'none'.")
  }
  
  # find the distances from the myx to all training points
  diff = Xs - matrix(myx_s, nrow = n, ncol = p, byrow=T)
  
  # ---------------- Apply Weights (optional) -----------------
  if (!is.null(weight)) {
    if (length(weight) == p) {
      # per-feature weights -> scale columns
      diff = sweep(diff, 2, weight, `*`)
    } else if (length(weight) == n) {
      # per-observation weights -> scale rows
      diff = diff * matrix(weight, nrow = n, ncol = p, byrow=F)
    }
  }
  
  # ----------------- Distance Metrics -----------------
  if(type == 1) {
    dist = rowSums(abs(diff))
  } else if(type == 2) {
    dist = sqrt(rowSums(diff^2))
  } else if(type == 'infty') {
    dist = apply(abs(diff), 1, max)
  } else {
    dist = (rowSums(abs(diff))^type)^(1/type)
  } 
  
  # find the k-nearest neighbors
  Y = factor(Y)
  idx = order(dist)[1:k]
  ylabel = Y[idx]
  ydist = dist[idx]
  
  # majority vote
  tab = table(ylabel)
  major_vote = names(tab)[tab == max(tab)][1]
  
  # distribution table
  distri_table = data.frame(class = names(tab),
                            count = as.integer(tab),
                            probs = as.integer(tab)/k)
  
  result = list('k_nearest_labels'   = ylabel,
                'k_nearest_dist'     = ydist,
                'distribution_table' = distri_table,
                'prob'               = max(as.integer(tab)/k),
                'yhat'               = major_vote)
  
  return(result)
}

################################################################################
### KNN accuracy o for test/train split
myknnacc = function(Xtrain, Ytrain, Xtest, Ytest, k, type=2, ztype="z-score", weight=NULL) {
  ntest = nrow(Xtest)
  myclass = c()
  myprobs = c()
  for (i in 1:ntest) {
    myknn = knn(Xtrain, Ytrain, Xtest[i,], k, type, ztype, weight)
    myclass = c(myclass, myknn$yhat)
    myprobs = c(myprobs, myknn$prob)
  }
  acc = mean(myclass == Ytest)
  
  result = list('yhat'     = myclass,
                'probs'    = myprobs,
                'accuracy' = acc)
  return(result)
}
```

## 2. K-Means Clustering (KMC)

```{R}
kmc = function(X, k, init='Forgy', iter='Lloyd', seed) {
  # X is n x p matrix with n observation and p features
  # k is the number of clusters we want
  set.seed(seed)
  n = nrow(X)
  p = ncol(X)
  
  # check whether k here is a valid parameter
  if(k < 1 || k > n) {
    stop("invalid k value: k should be = or < than the sample size n")
  }
  
  # ----------------- Initialization of Centroids ------------------------
  if(init == 'Forgy') {
    # choose k random existing points as initial centers
    centroids = X[sample.int(n, k, replace=F), , drop=F]
  } else if (init == 'Bounding-Box') {
    # find the bounding box for the entire dataset and uniformly random select
    # smaple k centroids inside the box
    mins = apply(X, 2, min)
    maxs = apply(X, 2, max)
    centroids = matrix(NA, nrow = k, ncol = p)
    for(j in 1:p) {
      centroids[,j] = runif(k, mins[j], maxs[j])
    }
  } else if (init == 'Random-Partition') {
    # randomly assign labels (1, 2, ..., k) to all points, then take the mean 
    # of each label as the initial centroids
    init = sample.int(k, n, replace=T)
    centroids = matrix(NA, nrow = k, ncol = p)
    for (j in 1:k) {
      idx = (init == j)
      centroids[j,] = colMeans(X[idx, , drop=F])
    }
  } else {
    stop("Unknown initialization metric. Use 'Forgy', 'Bounding-Box', or 'Random-Partition'.")
  }
  
  initial_centroids = centroids
  
  # -------------------------- Iterations ---------------------------------
  # initialize the clusters vector, which should contains only labels
  if (iter == 'Lloyd') {
    clusters = rep(0, n)
    iter = 0
    repeat {
      iter = iter + 1
      dist = matrix(0, nrow = n, ncol = k)
  
      # compute the 2-norm distance between each points and the k centers
      for (j in 1:k) {
        diff = X - matrix(centroids[j, ], nrow = n, ncol = p, byrow=T)
        dist[, j] = sqrt(rowSums(diff^2))
      }
      
      # assign cluster label to every other points
      new_clusters = apply(dist, 1, which.min)
      change = sum(new_clusters != clusters)
      clusters = new_clusters
      
      # move centers to the mean of assigned points
      for (j in 1:k) {
        idx = (clusters == j)
        if (any(idx)) {
          centroids[j, ] = colMeans(X[idx, , drop=F])
        }
      }
      
      # convergence check: labels unchanged
      if (change == 0) break
    }
  } else if (iter == 'Mac-Queen') {
    clusters = rep(0, n)
    prev_clusters = rep(0, n)
    iter = 0
    repeat {
      iter = iter + 1
      counts = rep(0, k)
      prev_clusters = clusters
      
      for (i in 1:n) {
        # distance from Xi to each center
        diff = centroids - matrix(X[i,], nrow = k, ncol = p, byrow=T)
        dist = sqrt(rowSums(diff^2))
        
        # nearest cluster; j here is the temporary centroids
        j = which.min(dist)
        clusters[i] = j
        
        # Mac-Queen incremental updates of centroids
        counts[j] = counts[j] + 1
        centroids[j,] = centroids[j,] + (X[i,] - centroids[j,])/counts[j]
      }
      
      if (sum(prev_clusters != clusters) == 0) break
    }
  } else if (iter == 'hierarchy') {
    clusters = rep(0, n)
    iter = 1
    for(i in 1:n) {
      diff = centroids - matrix(X[i,], nrow = k, ncol = p, byrow=T)
      dist = sqrt(rowSums(diff^2))
      clusters[i] = which.min(dist)
    }
  } else {
    stop("Unknown iteration metric. Use 'Lloyd', 'Mac-Queen', or 'hierarchy'.")
  }
  
  # ------------------------- Output and Result -----------------------------
  # build a distribution table: how many points in each cluster
  tab = table(clusters)
  nj = as.numeric(tab)
  
  Xdf = data.frame(X)
  Xs = split(Xdf, clusters)
  
  SSC = numeric(k)
  for (j in 1:k) {
    idx = (clusters == j)
    if (any(idx)) {
      Xj = X[idx, , drop=F]
      diff = Xj - matrix(centroids[j, ], nrow=nrow(Xj), ncol=p, byrow=T)
      SSC[j] = sum(rowSums(diff^2))
    }
  }
  TSSC = sum(SSC)
  
  nearest_cluster = function(myx) {
    diff = centroids - matrix(myx, nrow = k, ncol = p, byrow=T)
    dist = sqrt(rowSums(diff^2))
    return(which.min(dist))
  }
  
  result = list('iteration_num'     = iter,
                'initial_centroids' = initial_centroids,
                'final_centroids'   = centroids,
                'cids'              = clusters,
                'nj'                = nj,
                'SSC'               = SSC,
                'aSSC'              = SSC/nj,
                'TSSC'              = TSSC,
                'aTSSC'             = TSSC/sum(nj),
                'nearest_cids'      = nearest_cluster)
  return(result)
}

```
